---
title: "How to use tidyr::pivot_wider"
---

rmarkdown::render_site()

```{css, echo = FALSE}
body {
  font-family: arial;
  font-size: 13pt;
  color: black;
}
```

<style> body {text-align: justify} </style>

<br>
 
## Goal: Convert data format from long to wide

There are times when one wants to widen/spread a column i,e., convert a column to have its unique values as separate columns, hence making the data format long to wide. One can easily achieve this by using the pivot_wider function from the tidyr package. Let me walk you through with an example:

```{r message=FALSE, warning=FALSE, results='asis', tidy=F}
library(tidyr)

metric1 <- c(15.3, 14.7, 15.2, 13.8, 15.4, 14.8, 14.9, 15.1, 15.9)
id <- c("A01", "A01", "A01", "B01", "B01", "B01", "B02", "B02", "B02")
date <- as.Date(rep(c("2022-01-01", "2022-01-02", "2022-01-03"), 3))

df <- data.frame(metric1, id, date)

df_spread <- df %>% tidyr::pivot_wider(names_from = date, values_from = metric1)

```

```{r message=FALSE, warning=FALSE, results='asis', tidy=F}
# Before piot_wider()
knitr::kable(df, align = "c", caption = "<center>Table 1</center>") %>% 
  kableExtra::kable_styling(position = "center", full_width = F) 

# After piot_wider()
knitr::kable(df_spread, align = "c", caption = "<center>Table 2</center>") %>% 
  kableExtra::kable_styling(position = "center", full_width = F) 

```
